Gradual Soundness: Lessons from Static Python

### 2022-01-13 pj-7.1/paper.tex
pj: progress on sec 2 edits

This paper begins with an informal description of \SP{} in two parts.
First, we present a user-oriented tour of the language~(\cref{s:tour}).
Second, we present the key ingredients of the runtime system, Cinder,
that supports \SP{}~(\cref{s:runtime}).
\Cref{s:model} uses a small formal model to introduce our Redex mechanization
and to convince readers that \SP{} is based on a sound core.
\Cref{s:impl} notes important aspects of \SP{} that are not covered by the model.
\Cref{s:eval} evaluates \SP{}; it reports our experience with the language in
production and on public microbenchmarks.
The paper concludes with related work~(\cref{s:related}),
future work~(\cref{s:future}), and a final summary~(\cref{s:conclusion}).

@@

One important feature of \SP{} is that class hierarchies can
mix typed and untyped code.
An untyped class may inherit from a typed one and vice-versa.
The point of this feature is to let programmers add types to a single class
independently of its ancestors and children.

@@

\subsection{Unboxed Primitive}
\label{s:c-types}

@@

For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types
that describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
Cinder implements these types with unboxed C values, which are much simpler and cheaper
than their Python counterparts.
\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives efficiently.
These are not themselves primitive.

Primitives are carefully limited by the type system.
Neither a module-level nor a closure-level variable may have a primitive
type---because untyped code can mutate such variables.
Primitive types are incompatible with the dynamic type,
and there are no implicit upcasts from primitives
to a matching Python type.
Even a conjunction disallows mixing; for example, the \code{and} operator
requires either two Python booleans or two primitive booleans.
The only way for Python values and primitive values to interact
is through dedicated conversion functions.
Cinder handles conversions at the boundaries between \SP{} code
and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
Within typed code, programmers must satisfy the type checker with
appropriate conversions.

@@

>>>>

This paper begins with an informal description of \SP{} in two parts.
First, we present a user-oriented summary of the language~(\cref{s:tour}).
Second, we present the key ingredients of the runtime system, Cinder,
that supports \SP{}~(\cref{s:runtime}).
\Cref{s:model} uses a small formal model to introduce our Redex mechanization
and to convince readers that \SP{} is based on a sound core.
\Cref{s:impl} notes important aspects of \SP{} that are not covered by the model.
\Cref{s:eval} evaluates \SP{}; it reports our experience with the language in
production and on public microbenchmarks.
The paper concludes with related work~(\cref{s:related}),
future work~(\cref{s:future}), and a final summary~(\cref{s:conclusion}).

@@

One important feature of \SP{} is that class hierarchies can
mix typed and untyped code.
An untyped class can inherit from a typed one and vice-versa.
The point of this feature is to let programmers add types to a single class
independently of its ancestors and children.

@@

\subsection{Progressive Primitive Types}
\label{s:c-types}

@@

For performance-critical applications, \SP{} includes a set of primitive types
that describe booleans and sized numbers.
Example types include \code{int64}, \code{uint64}, \code{double}.
\SP{} also includes two special datatypes, \code{Array} and \code{Vector}, that store primitives efficiently.

The purpose of these types is to enable unboxed values and arithmetic at
runtime.
Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
refinement over the semantics of basic Python types.
Primitive types cannot be implicitly cast to non-primitive types,
including the dynamic type (another violation of the static gradual guarantee).
Operations such as \code{+} do not accept a mix of primitive and non-primitive data.
These restrictions give programmers a static separation between fast-running
primitives and standard data.

% Neither a module-level nor a closure-level variable may have a primitive
% type---because untyped code can mutate such variables.

@@

%% 2022-01-12
%% Cinder implements primitives with unboxed C values, which are much simpler and cheaper
%% than their Python counterparts.
%%
%% Cinder handles conversions at the boundaries between \SP{} code
%% and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
%% Within typed code, programmers must satisfy the type checker with
%% appropriate conversions.
